---
title: "Short Course on R Tools: Introduction to Rcpp"
output:
  html_document:
    echo: true
    warning: false
    message: false
    error: false
editor_options: 
  markdown: 
    wrap: 72
---

## Outline {auto-animate="true"}

-   Motivation & Introduction
-   Getting Started with `evalCpp()` & `cppFunction()`
-   Using `sourceCpp()`
-   Data Types & Conversions
-   Standard Template Library (STL)
-   Using Rcpp in Packages
-   Advanced Topics & Resources

# Motivation & Introduction

-   **Performance Bottlenecks**: loops, recursion, complex operations

-   **Rcpp**: smooth bridge between R and C++

-   Do things you could not do before

-   **Advantages**:

    -   Low overhead function calls
    -   Access to STL (data structures & algorithms)
    -   Cleaner, maintainable code

## Basic Usage: `evalCpp()`

-   `evalCpp()` evaluates a single C++ expression. Includes and
    dependencies can be declared.
-   This allows us to quickly check C++ constructs.

```{r code-line-numbers="1|3-4|6-7"}
library(Rcpp)

evalCpp("2 + 2")  # simple test
#> [1] 4

evalCpp("std::numeric_limits<double>::max()")
#> [1] 1.79769e+308
```

# Simple Example

-   R Version of “is this number odd or even”

```{r code-line-numbers="1-4|5-6"}
isOdd_r <- function(num = 10L) {
  result = (num %% 2L == 1L)
  return(result)
}
c(isOdd_r(42L), isOdd_r(43L))
#> [1] FALSE TRUE
```

-   C++ Version of “is this number odd or even”

``` cpp
bool isOdd_cpp(int num = 10) {
  bool result = (num % 2 == 1);
  return result;
}
```

## Simple Example – Using `cppFunction()`

-   Rcpp Version of “is this number odd or even”

```{r code-line-numbers="1,5|2-4|6-7"}
Rcpp::cppFunction("
bool isOdd_cpp(int num = 10) {
  bool result = (num % 2 == 1);
  return result;
}")
c(isOdd_cpp(42L), isOdd_cpp(43L))
#> [1] FALSE TRUE
```

-   Use `cppFunction()` in R console or script
-   Fast prototyping & testing
-   Good for small functions\

# Second Example: VAR(1)

Let’s consider a simple possible $\mathrm{VAR}(1)$ system of $k$
variables.

For $k = 2$:

$$
X_t = X_{t-1} B + E_t
$$

where $X_t$ is a row vector of length 2, $B$ is a $2\times 2$ matrix,
and $E_t$ is a row of the error matrix of 2 columns.

## Second Example: VAR(1)

In R code, given both the coefficient and error matrices:

```{r}
rSim <- function(B, E) {
  X <- matrix(0, nrow(E), ncol(E))
  for (r in 2:nrow(E)) {
    X[r, ] = X[r-1, ] %*% B + E[r, ]
  }
  return(X)
}
```

In C++ (with Armadillo):

``` cpp
arma::mat cppSim(arma::mat B, arma::mat E) {
  int m = E.n_rows, n = E.n_cols;
  arma::mat X(m, n);
  X.row(0) = arma::zeros<arma::mat>(1, n);
  for (int r = 1; r < m; r++) {
    X.row(r) = X.row(r-1) * B + E.row(r);
  }
  return X;
}
```

## Second Example: VAR(1)

```{r code-line-numbers="1,9|1-9|11-12|13-17"}
Rcpp::cppFunction('arma::mat cppSim(arma::mat B, arma::mat E) {
  int m = E.n_rows, n = E.n_cols;
  arma::mat X(m, n);
  X.row(0) = arma::zeros<arma::mat>(1, n);
  for (int r = 1; r < m; r++) {
    X.row(r) = X.row(r-1) * B + E.row(r);
  }
  return X;
}', depends="RcppArmadillo")

a <- matrix(c(0.5, 0.1, 0.1, 0.5), nrow = 2)
e <- matrix(rnorm(10000), ncol = 2)
rbenchmark::benchmark(cppSim(a, e), rSim(a, e), order="relative")[, 1:4]

#>    test           replications elapsed relative  
#> 1  cppSim(a, e)  100         0.010   1.0      
#> 2  rSim(a, e)    100         0.728  72.8     
```

# Growth of Rcpp

-   Sometimes speed is not the only reason
    -   Easy access to C/C++ libraries
    -   C & C++ provide numerous libraries + APIs
    -   Rcpp eases data transfer\
-   Rcpp is currently used by
    -   3,053 CRAN packages
    -   250+ BioConductor packages
    -   many GitHub projects\

## Pagerank

```{r code-line-numbers="1-3|5-7|9-10"}
# remotes::install_github("andrie/pagerank")
library(pagerank)

cran <- "https://cloud.r-project.org"
pr <- compute_pagerank(cran)
round(100 * pr[1:5], 3)

#>   Rcpp    ggplot2     dplyr      MASS   magrittr  
#> 2.744     1.516       1.274    1.122    0.814  
```

## Percentage of Compiled Packages

```{r code-line-numbers="1-2|4-6|8-10|12-13"}
db <- tools::CRAN_package_db()       # R ≥ 3.4.0
db <- db[!duplicated(db[, 1]), ]

nTot <- nrow(db)
nRcpp <- length(tools::dependsOnPkgs("Rcpp", recursive = FALSE, installed = db))
nCompiled <- table(db[, "NeedsCompilation"])[["yes"]]

propRcpp <- nRcpp / nCompiled * 100
data.frame(
  tot = nTot, 
  totRcpp = nRcpp, 
  totCompiled = nCompiled,
  RcppPctOfCompiled = propRcpp
)

#>    tot totRcpp totCompiled RcppPctOfCompiled  
#> 1 22501    3053        4981          61.29  
```

# Rcpp in RStudio – Using `sourceCpp()`

-   Create standalone `.cpp` files:

``` cpp
// file.cpp
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector timesTwo(NumericVector x) {
  return x * 2;
}
```

-   Compile with:

```{r code-line-numbers="1-2|4-5"}
sourceCpp('file.cpp')
timesTwo(42)
#> 84
```

## Using `sourceCpp()`

-   We defined a C++ function
-   It operates on a numeric vector
-   Rcpp creates and registers a wrapper
-   Compiles, links, and loads it
-   Function available in R by its C++ name
-   Benefits: syntax highlighting, easier debugging\

# Another Example: Focus on Speed

Consider the Fibonacci:

$$
f(n) =
\begin{cases}
n, & n < 2, \\
f(n-1) + f(n-2), & n \ge 2.
\end{cases}
$$

R implementation:

```{r code-line-numbers="1,4|2|3|5-7"}
f <- function(n) {
  if (n < 2) return(n)
  f(n - 1) + f(n - 2)
}
sapply(0:10, f)
#> 0 1 1 2 3 5 8 13 21 34 55
```

## Timing R Implementation

```{r code-line-numbers="1-2|4-7"}
library(rbenchmark)
benchmark(f(15), f(20), f(25))[, 1:4]

#> test   replications elapsed relative  
#> f(15) 100         0.06    1.0      
#> f(20) 100         0.75   12.5      
#> f(25) 100         7.92  132.0     
```

## Another Example: C++ Implementation

``` cpp
int g(int n) {
  if (n < 2) return(n);
  return(g(n - 1) + g(n - 2));
}

Rcpp::cppFunction('
int g(int n) {
  if (n < 2) return n;
  return g(n - 1) + g(n - 2);
}')
sapply(0:10, g)
#> 0 1 1 2 3 5 8 13 21 34 55
```

## Comparing Timing

```{r code-line-numbers="1-2|4-6"}
Rcpp::cppFunction('
int g(int n) {
  if (n < 2) return n;
  return g(n - 1) + g(n - 2);
}')

library(rbenchmark)
benchmark(f(25), g(25))[, 1:4]

#>    test replications elapsed relative  
#> 1 f(25) 100         8.02    267.33  
#> 2 g(25) 100         0.03      1.00  
```

-   Speed gains of several orders of magnitude.

# Data Types & Conversions

-   **R vectors ↔ C++ classes**: `NumericVector`, `IntegerVector`,
    `CharacterVector`, `LogicalVector`
-   **Scalars**: `double`, `int`, `String`, `bool`
-   **Key methods**: `.size()`, `.begin()`, `.end()`
-   Constructors: `NumericVector out(n)`

## Types

R type mapping in C++:

```{r code-line-numbers="1-4|5-6"}
library(Rcpp)
cppFunction("NumericVector logabs(NumericVector x) { return log(abs(x)); }")
logabs(seq(-5, 5, by = 2))
#> 1.609438 1.098612 0.000000 0.000000 1.098612 1.609438
```

-   Vectorized C++!

## STL Type Mapping

``` cpp
// logabs2.cpp
#include <Rcpp.h>
using namespace Rcpp;

inline double f(double x) { return ::log(::fabs(x)); }

// [[Rcpp::export]]
std::vector<double> logabs2(std::vector<double> x) {
  std::transform(x.begin(), x.end(), x.begin(), f);
  return x;
}
```

```{r}
library(Rcpp)
sourceCpp("code/logabs2.cpp")
logabs2(seq(-5, 5, by = 2))
#> 1.609438 1.098612 0.000000 0.000000 1.098612 1.609438
```
