---
title: "Short Course on R Tools: Object-Oriented Programming in R"
output:
  html_document:
    echo: true
    warning: false
    message: false
    error: false
---

# Outline

-   **OOP Paradigms in R**
-   **Base Types & Functional OOP**
-   **S3 Classes & Methods**
-   **S4 Classes & Methods**
-   **R6 Classes & Reference Semantics**
-   **Trade-offs & Best Practices**
-   **Hands-on Exercises**

------------------------------------------------------------------------

## OOP Paradigms in R

> **Functional OOP**: functions return objects, operate on data **Encapsulated OOP**: objects carry data + methods **Generics & Method Dispatch**: generic functions choose methods based on class

------------------------------------------------------------------------

## Example: Functional OOP

```{r}
# (1) Plotting base graphics for demonstration of functional OOP:
par(mfrow = c(1, 3))

# Scatterplot of Sepal.Length
plot(iris$Sepal.Length, main = "Sepal Length", xlab = "", ylab = "")

# Plot of Species (factor)
plot(iris$Species, main = "Species", xlab = "", ylab = "")

# Linear model example (commented out; remove comments to run)
# plot(lm(Sepal.Length ~ Sepal.Width, data = iris), main = "LM Diagnostics")
```

------------------------------------------------------------------------

## Encapsulated OOP Example (using mlr3)

```{r}
library(mlr3)

# Create a classification task (encapsulates data + metadata)
task    <- TaskClassif$new(id = "iris_task", backend = iris, target = "Species")
learner <- lrn("classif.rpart", predict_type = "prob")

# Real use of object$method(args)
learner$train(task)
prediction <- learner$predict(task)

# Inspect first few (and last few) predicted probabilities
head(prediction$prob, 3)
tail(prediction$prob, 3)
```

------------------------------------------------------------------------

# OOP in R: Overview of Systems

-   **S3**: Informal, functional OOP (lightweight class attribute, UseMethod dispatch)
-   **S4**: Formal, slots, strict validation, implemented in the *methods* package
-   **RC** (Reference Classes): built-in reference semantics (less common)
-   **R6**: Modern encapsulated OOP with mutable fields and methods
-   **S7**: New unified OOP system (multiple dispatch, type safety, S3/S4 compatibility)

------------------------------------------------------------------------

## Detecting Object Type

```{r}
library(sloop)

otype(1:10)
#> [1] "base"

otype(mtcars)
#> [1] "S3"

mle_obj <- stats4::mle(function(x = 1) (x - 2)^2)
otype(mle_obj)
#> [1] "S4"
```

```{r}
# Base vs OO objects
is.object(1:10)
#> [1] FALSE

is.object(mtcars)
#> [1] TRUE

# Inspecting the 'class' attribute
attr(1:10, "class")
#> NULL

attr(mtcars, "class")
#> [1] "data.frame"
```

------------------------------------------------------------------------

# Base Types in R

*Every object in R has a base type. Below are some examples of base types:*

```{r}
typeof(NULL)         # "NULL"
typeof(1L)           # "integer"
typeof(1i)           # "complex"
typeof(mean)         # "closure"
typeof(`[`)          # "special"
typeof(sum)          # "builtin"
typeof(globalenv())  # "environment"

# An S4 object also registers as a distinct type
mle_obj <- stats4::mle(function(x = 1) (x - 2)^2)
typeof(mle_obj)      # "S4"
```

------------------------------------------------------------------------

# Informal OOP in R (S3-like)

-   Build an object as a plain list
-   Assign a `class` attribute
-   Define print/summary/etc. methods by naming functions as `generic.classname`

```{r}
# Create a 'weather_report' object (simple list)
weather <- list(
  location = list(city = "Seattle", country = "USA", coordinates = c(lat = 47.6062, lon = -122.3321)),
  date = as.Date("2025-05-30"),
  summary = "Cloudy with light rain",
  temperature = list(morning = 57, afternoon = 63, evening = 60, units = "F"),
  alerts = c("Flood watch", "Wind advisory")
)

# Assign S3 class
class(weather) <- "weather_report"

# Define a print method for the 'weather_report' class
print.weather_report <- function(x) {
  cat("🌤️  Weather Report
")
  cat("-------------------------
")
  cat("📍 Location: ", x$location$city, ", ", x$location$country, "
", sep = "")
  cat("🗓️  Date: ", format(x$date, "%B %d, %Y"), "
", sep = "")
  cat("📋 Summary: ", x$summary, "
", sep = "")
  cat("🌡 Temperature (", x$temperature$units, "):
", sep = "")
  cat("    Morning:  ", x$temperature$morning, "
", sep = "")
  cat("    Afternoon:  ", x$temperature$afternoon, "
", sep = "")
  cat("    Evening:  ", x$temperature$evening, "
", sep = "")
  if (length(x$alerts) > 0) {
    cat("⚠️ Alerts: ", paste(x$alerts, collapse = "; "), "
", sep = "")
  } else {
    cat("✅ No active alerts.
")
  }
  invisible(x)
}

# Call the custom print method
print(weather)
```

------------------------------------------------------------------------

# S3: Informal Classes and Dispatch

-   **Class Definition**: assign class attribute
-   **Method Definition**: name functions as `generic.classname`

```{r}
# Create an S3 object in one line
x <- structure(list(a = 1, b = 2), class = "myclass")

# Define a custom mean method for 'myclass'
mean.myclass <- function(x, ...) {
  print("custom mean for myclass")
}

# Test the new method
mean(x)
```

## Inspecting S3 Details

```{r}
# Example: factor is an S3 class
f <- factor(c("a", "b", "b"))
attributes(f)

# Unclassing drops the 'factor' class
unclass(f)

# Time object (POSIXlt) is also S3
time <- strptime(c("2017-01-01", "2020-05-04 03:21"), "%Y-%m-%d")
str(time)

# Drop the class attribute and inspect underlying structure
class(time) <- NULL
str(time)
```

------------------------------------------------------------------------

# S3: Common Constructor Pattern

1.  **Low-level constructor**: `new_myclass()`
2.  **Validator**: `validate_myclass()`
3.  **User-facing helper**: `myclass()`

```{r}
## 1. Low-level constructor: assume inputs are valid
new_myclass <- function(values = numeric(), unit = character(1)) {
  stopifnot(is.numeric(values), is.character(unit), length(unit) == 1)
  structure(
    list(values = values, unit = unit),
    class = "myclass"
  )
}

## 2. Validator: check consistency
validate_myclass <- function(x) {
  if (!inherits(x, "myclass")) {
    stop("`x` must be a 'myclass' object.", call. = FALSE)
  }
  if (!is.numeric(x$values)) {
    stop("`values` must be numeric.", call. = FALSE)
  }
  if (any(is.na(x$values))) {
    stop("`values` must not contain NA.", call. = FALSE)
  }
  if (!is.character(x$unit) || length(x$unit) != 1) {
    stop("`unit` must be a single character string.", call. = FALSE)
  }
  TRUE
}

## 3. User-facing helper: build & validate
myclass <- function(values, unit = "unitless") {
  if (!is.numeric(values)) {
    values <- as.numeric(values)
    if (any(is.na(values))) {
      stop("`values` could not be coerced to numeric.", call. = FALSE)
    }
  }
  if (!is.character(unit) || length(unit) != 1) {
    stop("`unit` must be a single character string.", call. = FALSE)
  }
  obj <- new_myclass(values = values, unit = unit)
  validate_myclass(obj)
  obj
}

## Example usage
x <- myclass(c(1.1, 2.2, 3.3), unit = "kg")
print(x)

# This will error due to NA
# myclass(c(1, NA, 3))
```

------------------------------------------------------------------------

# S3: Generics and Method Dispatch

-   Use `UseMethod("generic")` inside generic definitions
-   Methods must match the generic’s arguments (unless generic uses `...`)

```{r}
library(sloop)

# View generic function signature for 'print'
show(print)
ftype(print)         # "S3" "generic"
ftype(print.factor)  # "S3" "method"

# Show dispatch path for printing a factor
f <- ordered(c("a", "b", "b"))
s3_dispatch(print(f))

# Show dispatch path for printing an unclassed factor
s3_dispatch(print(unclass(f)))
```

------------------------------------------------------------------------

# S3: Finding Methods

```{r}
# List all methods for generic 'mean'
s3_methods_generic("mean")

# List all generics that have methods for class 'funts'
s3_methods_class("funts")
```

------------------------------------------------------------------------

# S3: Inheritance and `NextMethod()`

```{r}
# Example: define a 'secret' class that masks its values
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(as.character(x))))
  invisible(x)
}

x <- new_secret(c(15, 1, 456))
x
#> [1] "xx"  "x"   "xxx"

# Default subsetting drops the class
x[1]
#> [1] 15

# Provide a `[.secret` method to preserve the class
`[.secret` <- function(x, i) {
  new_secret(NextMethod())
}

x[1]
#> [1] "xx"
```

------------------------------------------------------------------------

# S4: Formal Classes

```{r}
# Define a new S4 class "Person" with slots
setClass("Person",
  slots = list(
    name = "character",
    age  = "numeric"
  )
)

# Create an instance of "Person"
subj1 <- new("Person", name = "Adam", age = 30)
subj1

# Access slots via "@" or slot()
subj1@name
slot(subj1, "age")
```

## S4: Generics & Methods

```{r}
# Define a generic function "greet"
setGeneric("greet", function(x) standardGeneric("greet"))

# Define a method for class "Person"
setMethod("greet", "Person", function(x) {
  paste("Hello, my name is", x@name)
})

# Call the new method
greet(subj1)
#> [1] "Hello, my name is Adam"
```

## S4: Inheritance & Virtual Classes

```{r}
# Define a subclass "Employee" that contains "Person"
setClass("Employee",
  contains = "Person",
  slots = list(salary = "numeric")
)

# Create an instance of "Employee"
emp1 <- new("Employee", subj1, salary = 75000)
emp1

# Inspect slots
emp1@name
emp1@age
emp1@salary
```

------------------------------------------------------------------------

# R6: Reference Classes (Encapsulated OOP)

-   Mutable fields
-   Methods inside the class definition
-   Use `$new()` to instantiate

```{r}
library(R6)

# Define an R6 class "PersonR6"
PersonR6 <- R6Class(
  "PersonR6",
  public = list(
    name = NULL,
    initialize = function(name) {
      self$name <- name
    },
    greet = function() {
      cat("Hi, I'm", self$name, "
")
    }
  )
)

# Instantiate and use
p <- PersonR6$new("Bob")
p$greet()
#> Hi, I'm Bob
```

------------------------------------------------------------------------

# S7: The New OOP System {auto-animate="true"}

-   Designed by the R Consortium to unify S3+S4 and fix issues.
-   **Key features**:
    -   Multiple dispatch
    -   Formal class definitions
    -   Compatibility with S3/S4
    -   Faster than S4, simpler than R6

```{r}
library(S7)
Circle <- new_class("Circle", properties = list(radius = class_double))
circle1 <- Circle(radius = 5)
typeof(circle1)
#> [1] "object"
circle1
#> <Circle>
#>  @ radius: num 5
```
